TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains

### 2019-09-08 manuscript/manuscript.Rmd
Replace manuscript, adjust for latest paths

---
bibliography: [refs/networks.bib,refs/brain.bib,refs/enrichment.bib,refs/3D_refs.bib]
csl: styles.ref/genomebiology.csl
output:
  pdf_document: default
  word_document:
    reference_docx: styles.doc/Arial_11_single_space_normal_margins.docx
  html_document: default
editor_options: 
  chunk_output_type: console
---

@@

**Results.** Using STRING and BioGrid PPI databases, we compared the coherence of 133 phenotype-associated gene sets from GWAScatalog against size-matched KEGG pathways (the standard of high coherence) and random networks (the lower limit of coherence). We observed a range of coherence estimates for each category of phenotypes. Metabolic traits and diseases were the most coherent, while psychiatric diseases and intelligence-related traits were the least coherent.

**Conclusions.** We present a general-purpose method for estimating and comparing the coherence of phenotype-associated gene networks that accounts for the network size and shape differences. Our results highlight the gaps in of our current knowledge of the genetics and molecular mechanisms of complext phenotypes and suggest priorities for future work.

@@

We created networks of phenotype-associated genes using PPI information from STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa] databases (see Methods). We opted for the use of three types of PPI data. First, STRING data filtered by the interaction confidence score above 500 (middle of the bimodal distribution of the score, 0-1000 range, referred hereafter as “STRING filtered”) was used as the primary source of curated PPI data. This filtering step ensured that only high-confidence PPIs were selected. The advantage of using filtered data is the reliance on high-quality PPIs. The disadvantage is that the number of PPIs may be insufficient for forming phenotype-associated networks; consequently, phenotypes with genes without PPIs are removed from the analysis (see Methods). Second, unfiltered STRING data were used to maximize the use of PPI information at the expense of potential noise (referred hereafter as “STRING full”). The advantage of using unfiltered data is that more phenotype-associated genes will have PPI information and, hence, can be analyzed. The disadvantage, however, is that the results may be less reliable due to the presence of noisy PPIs. Third, BioGrid PPI data, which has been curated to keep only high-confidence interactions, were used. The use of different data sources (STRING and BioGrid) and filtering (full or filtered STRING data) was intended to increase the generalizability of our conclusions.

**Table 1. Summary statistics of the analyzed categories.** Average coherence estimates are shown. "NA" values indicate that a phenotype was not considered due to the low number of genes with PPIs.

| Category               | Total Phenotypes | Number of SNPs | Number of Genes | Biogrid Coherence | String Coherence | String Filtered Coherence |
|------------------------|:----------------:|:--------------:|:---------------:|:-----------------:|:----------------:|:-------------------------:|
| Disease                |                  |                |                 |                   |                  |                           |
| Metabolic disease      |         5        |       345      |       280       |        0.86       |       0.79       |           0.933           |
| Neurologic             |         8        |       382      |       343       |        0.89       |       0.511      |           0.861           |
| Eye disease            |         8        |       270      |       242       |        0.75       |       0.536      |           0.856           |
| Autoimmune             |        20        |      1415      |       1212      |       0.807       |       0.801      |           0.852           |
| Cancer                 |        18        |       713      |       610       |       0.863       |       0.643      |           0.838           |
| Cardiovascular disease |         9        |       336      |       289       |       0.705       |       0.584      |           0.832           |
| Psychiatric            |        11        |       678      |       611       |        0.83       |       0.537      |           0.817           |
| Trait                  |                  |                |                 |                   |                  |                           |
| Metabolic trait        |        10        |       790      |       646       |       0.906       |       0.776      |            0.93           |
| Eye trait              |         5        |       158      |       109       |         NA        |       0.674      |           0.903           |
| Anthropometric trait   |        19        |      2143      |       1822      |       0.793       |       0.601      |           0.841           |
| Cardiovascular trait   |        15        |       728      |       604       |       0.833       |       0.647      |           0.813           |
| Intelligence           |         5        |       236      |       198       |        0.78       |       0.312      |           0.757           |

@@

+-------+
| ![](Figures/Size_Slope_Plot.png)      |
| **Figure 2. Network size is inversely associated with coherence.** Slopes of internal vs. external degree distributions (aka coherence) for KEGG and random networks are plotted against network size.   |
+-------+

@@

Phenotype-associated gene lists were obtained from the NHGRI-EBI catalog of genome-wide association studies (GWAS catalog, `gwascat` v.2.14.0 R package) [@Hindorff:2009aa; @MacArthur:2017aa]. "MAPPED_GENE" column was used to extract phenotype-associated genes. To organize phenotypes into categories, Experimental Factor Ontology (EFO) IDs were extracted from the "MAPPED_TRAIT_URI" column. Categories were mapped to EFO IDs using the `ontoCAT` v.1.12.0 R package and the EFO database (accessed 12/04/2018). In the case of a phenotype being mapped into two categories, the most representative category was manually assigned. Non-canonical gene names were converted to common gene symbols using `alias2Symbol` function from the `limma` R package. Gene names which could not be mapped to gene symbols were excluded from the analysis.

@@

Phenotype-associated gene networks were formed by mapping phenotype-associated genes to PPIs. Genes without PPI annotations (e.g., C5orf4, LOC100507462), antisense transcripts (e.g., CDKN2B-AS), readthrough (e.g., NPHP3-ACAD11), non-protein-coding transcripts (e.g., LINC00478) were excluded. Genes that do not have PPIs with each other but interact with other genes in the global PPI network were kept. Phenotypes with less than five genes were omitted. Additionally, phenotypes with genes having identical internal degrees (hence, no degree distribution) were filtered due to the inability to derive a fit through the internal vs. external degree distributions (see below). These precautions were set to avoid situations where a small number of genes in a network can cause instability in the results. Importantly, the number of phenotype-associated genes having PPI annotations depends on the content of a PPI database; hence, some phenotypes may be filtered when using one but not the other PPI database to build phenotype-associated networks.

@@

The **internal degree distribution** was defined as the number of interactions between genes of interest, or, more precisely, their interacting protein products. The **external degree distribution** was defined as the number of interactions between genes of interest and all other genes in the PPI network.

To compare phenotype-specific degree distributions, the internal vs. external degree distributions of phenotype-specific genes were plotted on a scatterplot. The square root of the degree was taken for more informative visualization. For each phenotype, the internal vs. external degree distribution plot was fit with a regression line through the origin to capture the slopes. Slopes were used as a measure of coherence with lower slopes being associated with higher coherence. Degree distributions of the network nodes were obtained using the `igraph` R package v. 1.0.1.

@@

The slopes of the internal vs. external degree distributions of the networks exhibiting high and random coherence (KEGG and random networks, respectively)  are dependent on the content of a PPI database and the network size. To compare coherences using a given PPI database, we normalized the slopes for the phenotype-associated networks using slopes formed by the degree distributions of similarly sized networks with high coherence ($\beta_{min}$ slope of KEGG networks) and random coherence ($\beta_{max}$ slope of random networks). For each phenotype $i$ with slope $\beta_i$, we calculate the normalized slope $\beta_{i\ norm} = \frac{\beta_{i} - \beta_{max}}{\beta_{min}- \beta_{max}}$, which gives a measure of coherence on the range of [0,1]. Higher values indicate a higher level of coherence similar to that of KEGG pathways. These values represent normalized coherence and can be compared across phenotypes and PPI databases.

@@

To test whether phenotype-specific coherence is significantly different than random, we used a permutation test. For phenotype $i$ with the gene network of size $g_i$, we define $\beta_i$ as the slope derived by regressing the external degree distribution on the internal degree distribution. We then generate $n=10000$ networks of randomly selected genes of the same size $g_i$ as the set for phenotype $i$. For each random network $j$, we identify its slope $\beta_{r}$ representing the internal vs. external degree distribution regression line of random coherence. We calculate the permutation p-value as $p_i = \frac{(\sum_{j=1}^nI({\beta_{rj} > \beta_i}) + 1}{n + 1}$ [@Phipson:2010aa]. Intuitively, this approach estimates whether the slope of the phenotype-specific network $\beta_i$ is consistently larger than those of random networks. All analyses and visualizations were performed in the R/Bioconductor computing environment v.3.4.0 [@gentleman2006bioinformatics].

@@

This work was supported by a Genetics and Human Agency Award from the Templeton Foundation to KSK, MR, CC, SB. The content is solely the responsibility of the authors and does not necessarily represent the official views of the Templeton Foundation.

@@

**Supplementary Figure S3. Coherence estimates of phenotypes.** Size of dots represents the level of normalized coherence (X-axis) for individual phenotypes (Y-axis). Missing entries indicate that for a given PPI database a phenotype-associated network could not be built.

>>>>

---
bibliography: [/Users/mdozmorov/Documents/Work/VCU_grants/1_Dozmorov/networks.bib,/Users/mdozmorov/Documents/Work/VCU_grants/1_Dozmorov/brain.bib,/Users/mdozmorov/Documents/Work/VCU_grants/1_Dozmorov/enrichment.bib,/Users/mdozmorov/Documents/Work/VCU_grants/1_Dozmorov/3D_refs.bib]
csl: /Users/mdozmorov/Documents/Work/presentations/Templates/grant_template/styles.ref/genomebiology.csl
output:
  word_document:
    reference_docx: /Users/mdozmorov/Documents/Work/presentations/Templates/manuscript_template/styles.doc/Arial_11_single_space_normal_margins.docx
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

@@

**Results.** Using STRING and BioGrid PPI databases, we compared the coherence of 133 phenotype-associated gene sets from GWAScatalog against size-matched KEGG pathways (with exemplary high coherence) and random networks (the lower limit of coherence). We observed a range of coherence estimates for each category of phenotypes. Metabolic traits and diseases were the most coherent, while psychiatric diseases and intelligence-related traits were the least coherent.

**Conclusions.** We present a general-purpose method for estimating and comparing the coherence of phenotype-associated gene networks that accounts for the network size and shape differences. Our results highlight the gaps in of our current knowledge of the genetics and molecular mechanisms of complext traits and suggest priorities for future work.

@@

We created networks of phenotype-associated genes using PPI information from STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa] databases (see Methods). We opted for the use of three types of PPI data. First, STRING data filtered by the interaction confidence score above 500 (middle of the bimodal distribution of the score, 0-1000 range, referred hereafter as “STRING filtered”) was used as the primary source of curated PPI data. This filtering step ensured that only high-confidence PPIs were selected. The advantage of using filtered data is the reliance on high-quality PPIs. The disadvantage is that the number of PPIs may be insufficient for forming phenotype-associated networks; consequently, phenotypes with genes without PPIs are removed from the analysis (see Methods). Second, unfiltered STRING data were used to maximize the use of PPI information at the expense of potential noise (referred hereafter as “STRING full”). The advantage of using unfiltered data is that more trait-associated genes will have PPI information and, hence, can be analyzed. The disadvantage, however, is that the results may be less reliable due to the presence of noisy PPIs. Third, BioGrid PPI data, which has been curated to keep only high-confidence interactions, were used. The use of different data sources (STRING and BioGrid) and filtering (full or filtered STRING data) was intended to increase the generalizability of our conclusions.

<!--Tables/table_1.csv-->
**Table 1. Summary statistics of the analyzed categories.** Average coherence estimates are shown. "NA" values indicate that a phenotype was not considered due to the low number of genes with PPIs.

| Category               | Total Phenotypes | Number of SNPs | Number of Genes | Biogrid | String | String Filtered |
|------------------------|:----------------:|:--------------:|:---------------:|:-------:|:------:|:---------------:|
| Disease                |                  |                |                 |         |        |                 |
| Metabolic disease      |         5        |       345      |       280       |  0.515  |  0.776 |      0.845      |
| Autoimmune             |        20        |      1415      |       1212      |  0.470  |  0.801 |      0.702      |
| Eye disease            |         8        |       270      |       242       |  0.300  |  0.509 |      0.654      |
| Neurologic             |         8        |       382      |       343       |  0.710  |  0.498 |      0.637      |
| Cancer                 |        18        |       713      |       610       |  0.548  |  0.658 |      0.604      |
| Cardiovascular disease |         9        |       336      |       289       |  0.130  |  0.577 |      0.602      |
| Psychiatric            |        11        |       678      |       611       |  0.560  |  0.528 |      0.583      |
| Trait                  |                  |                |                 |         |        |                 |
| Metabolic trait        |        10        |       790      |       646       |  0.766  |  0.779 |      0.842      |
| Eye trait              |         5        |       158      |       109       |    NA   |  0.678 |      0.698      |
| Anthropometric trait   |        19        |      2143      |       1822      |  0.600  |  0.624 |      0.678      |
| Cardiovascular trait   |        15        |       728      |       604       |  0.463  |  0.655 |      0.537      |
| Intelligence           |         5        |       236      |       198       |  0.310  |  0.334 |      0.373      |

@@

+-------+
| ![](Figures/Fig2_Size_Slope_Plot.png)      |
| **Figure 2. Network size is inversely associated with coherence.** Slopes of internal vs. external degree distributions (aka coherence) for KEGG and random networks are plotted against network size.   |
+-------+

@@

Trait-associated gene lists were obtained from the NHGRI-EBI catalog of genome-wide association studies (GWAS catalog, `gwascat` v.2.14.0 R package) [@Hindorff:2009aa; @MacArthur:2017aa]. "MAPPED_GENE" column was used to extract trait-associated genes. To organize traits into categories, Experimental Factor Ontology (EFO) IDs were extracted from the "MAPPED_TRAIT_URI" column. Categories were mapped to EFO IDs using the `ontoCAT` v.1.12.0 R package and the EFO database (accessed 12/04/2018). In the case of a trait being mapped into two categories, the most representative category was manually assigned. Non-canonical gene names were converted to common gene symbols using `alias2Symbol` function from the `limma` R package. Gene names which could not be mapped to gene symbols were excluded from the analysis.

@@

Trait-associated gene networks were formed by mapping trait-associated genes to PPIs. Genes without PPI annotations (e.g., C5orf4, LOC100507462), antisense transcripts (e.g., CDKN2B-AS), readthrough (e.g., NPHP3-ACAD11), non-protein-coding transcripts (e.g., LINC00478) were excluded. Genes that do not have PPIs with each other but interact with other genes in the global PPI network were kept. Traits with less than five genes were omitted. Additionally, traits with genes having identical internal degrees (hence, no degree distribution) were filtered due to the inability to derive a fit through the internal vs. external degree distributions (see below). These precautions were set to avoid situations where a small number of genes in a network can cause instability in the results. Importantly, the number of trait-associated genes having PPI annotations depends on the content of a PPI database; hence, some traits may be filtered when using one but not the other PPI database to build trait-associated networks.

@@

Trait-specific **internal degree distribution** was defined as the number of interactions between trait-associated genes, or, more precisely, their interacting protein products. As different traits have different numbers of associated genes forming networks of unknown configurations, the corresponding internal degree distribution was normalized to the expected interaction count for a network of that size (see Results).

Trait-specific **external degree distribution** was defined as the number of interactions between genes in a trait-associated network and all other genes in the PPI network. Being measured in the context of the entire PPI network of fixed size, the external degree distribution did not require normalization.

To compare trait-specific degree distributions, the normalized internal vs. external degree distributions were plotted on a scatterplot. The square root of the degree was taken for more informative representation. For each trait, the internal vs. external degree distribution plot was fit with a regression line through the origin to capture the slopes. Slopes were used as a measure of coherence with lower slopes being associated with higher coherence. Degree distributions of the network nodes were obtained using the `igraph` R package v. 1.0.1.

@@

The slopes of the internal vs. external degree distributions of the networks exhibiting high and random coherence (KEGG and random networks, respectively)  are dependent on the content of a PPI database. Thus, we normalized the slopes for the trait-associated networks using slopes formed by the degree distributions of similarly sized networks with high coherence ($\beta_{min}$ slope of KEGG networks) and random coherence ($\beta_{max}$ slope of random networks). For each trait $i$ with slope $\beta_i$, we calculate the normalized slope $\beta_{i\ norm} = \frac{\beta_{i} - \beta_{max}}{\beta_{min}- \beta_{max}}$, which gives a measure of coherence on the range of [0,1]. Higher values indicate a higher level of coherence similar to that of KEGG pathways. These values represent normalized coherence and can be compared across traits and PPI databases.

@@

To test whether each slope is significantly different than random, we used a permutation test. For trait $i$ with the gene network of size $g_i$, we define $\beta_i$ as the slope derived by regressing the external degree distribution on the internal degree distribution. We then generate $n=10000$ networks of randomly selected genes of the same size $g_i$ as the set for trait $i$. For each random network $j$, we identify its slope $\beta_{r}$ representing the internal vs. external degree distribution regression line of random coherence. We calculate the permutation p-value as $p_i = \frac{(\sum_{j=1}^nI({\beta_{rj} > \beta_i}) + 1}{n + 1}$ [@Phipson:2010aa]. Intuitively, this approach estimates whether the slope of the trait-specific network $\beta_i$ is consistently larger than those of random networks. All analyses and visualizations were performed in the R/Bioconductor computing environment v.3.4.0 [@gentleman2006bioinformatics].

@@

This work was supported by a Genetics and Human Agency Award from the Templeton Foundation to KSK, MR, CC, SB. The content is solely the responsibility of the authors and does not necessarily represent the official views of the Templeton Foundation. The authors would like to thank Spiro Stilianoudakis for help with Figure 1.

@@

**Supplementary Figure S3. Coherence estimates of traits.** Size of dots represents the level of normalized coherence (X-axis) for individual traits (Y-axis). Missing entries indicate that for a given PPI database a phenotype-associated network could not be built.
